Properties of winning Iterated Prisoner's Dilemma strategies.

### 2020-03-30 paper/main.tex
Revise introduction

\begin{abstract}
Researchers have explored the performance of Iterated Prisoner's Dilemma strategies
for decades: from the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated learning
structures such as neural networks, many new strategies have been introduced and tested
in a variety of tournaments and population dynamics. Typical results in the literature,
however, rely on performance against a small number of somewhat arbitrarily selected
strategies in a very small number of tournaments, casting doubt on the generalizability
of conclusions. We analyze a large collection of \numberofstrategies
strategies in \numberofalltournaments tournaments, present the top performing strategies across multiple
tournament types, and distill their salient features. The results show that there is not yet a single
strategy that performs well in diverse Iterated Prisoner's Dilemma scenarios.
Nevertheless there are several properties that heavily influence the best performing
strategies, refining the properties described by R. Axelrod in light of
recent and more diverse opponent populations. These are: be nice, be provocable and contrite,
be a little envious, be clever, and adapt to the environment, which includes the parameters
of the tournament (e.g. noise) and the population of opponents. More precisely,
we find that strategies perform best when their probability of cooperation
matches the total tournament population's aggregate cooperation probabilities,
or a proportion thereof in the case of noisy and probabilistically ending tournaments,
and that the manner in which a strategy achieves the ideal cooperation rate is crucial.
\end{abstract}

@@

The Iterated Prisoner's Dilemma (IPD) is a repeated two player game that models
behavioural interactions, and more specifically, interactions where
self-interest clashes with collective interest. At each turn of the game both
players, simultaneously and independently, decide between cooperation (C) and
defection (D) whilst having memory of their prior interactions. The payoffs for each
player, at each turn, is influenced by their own choice and the choice of the
other player. The payoffs of the game are generally defined by:

@@

where \(T > R > P > S\) and \(2R > T + S\). The most common values used in
the literature~\cite{Axelrod1981} are $R=3, P=1, T=5, S=0$. These values are also
used in this work.

Conceptualising strategies and understanding the best way of playing the game
has been of interest to the scientific community since the formulation of the
game in 1950~\cite{Flood1958}. Following the computer tournaments of Axelrod in the
1980's~\cite{Axelrod1980a, Axelrod1980b}, a strategy's performance in a round
robin computer tournament became a common evaluation technique for newly designed
strategies. Many tournaments have followed Axelrod's~\cite{Bendor1991,
Harper2017, Kendall2007, Stephens2002, Stewart2012} and today more than 200
strategies exist in the literature.

Axelrod performed two computer tournaments~\cite{Axelrod1980a,
Axelrod1980b} in the 80's. The
winner of both tournaments was the simple strategy Tit For Tat (TFT) which cooperated
on the first turn and then simply copied the previous action of its opponent.
Axelrod concluded that the strategy's robustness was due to four properties, which
he adapted in four suggestions on doing well in an IPD:

\begin{itemize}
    \item Do not be envious
    \item Be "nice"; Do not be the first to defect
    \item Reciprocate both cooperation and defection; Be provocable
    \item Do not be too clever
\end{itemize}

@@

Other works have built upon the limitations of the TFT, and others have shown
that suggestions made by Axelrod were incomplete.
In~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984} it was shown
that TFT suffered in environments with noise. This was
mainly due to the strategy's lack of generosity and contrition. The strategy immediately
punishes a defection and in a noisy environment it could lead to a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced and became the new
protagonists of the game. These include Nice and Forgiving~\cite{Bendor1991},
Pavlov (aka Win Stay Lose Shift) ~\cite{Nowak1993}, Generous Tit For Tat~\cite{Nowak1992}.
In~\cite{Press2012} a set of envious IPD strategies
were introduced called zero-determinant strategies (ZDs). By
forcing a linear relationship between stationary payoffs ZDs can ensure that they will
never receive less than their opponents. ZDs are indeed a set of mathematically
unique strategies and robust in pairwise interactions, however, their
superiority in tournament settings have been tested. In~\cite{Harper2017} a
series of clever strategies were introduced, and a tournament containing over
200 strategies was ran and none of the aforementioned strategies ranked in top
spots. Instead, the top ranked strategies were the set of clever trained
strategies based on lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017}, and finite state automata~\cite{Miller1996}.

Though only select pieces of work have been discussed, there is a broad number
of strategies in the literature, and new strategies and competitions are being
published every year~\cite{Glynatsi2019}. The question, however, still remains
the same: what is the best way to play the game?

Compared to other works, where a few selected strategies are evaluated
on a small number of tournaments, this manuscript evaluates the performance of \numberofstrategies
strategies in \numberofalltournaments tournaments. These tournaments come in a
number of variations including tournaments with noise, probabilistic match
length and both noise and probabilistic match length.
The later part of the paper evaluates
the impact of features on the performance of the strategies using modern
machine learning techniques. These features include measures regarding a
strategy's behaviour and measures regarding the tournaments. The outcomes of our
work reinforce the discussion started by Axelrod, and it concludes that the properties
of a successful strategy in the IPD are:

\begin{itemize}
    \item \st{Do not be envious} Be a little bit envious
    \item Be "nice"; Do not be the first to defect
    \item Reciprocate both cooperation and defection; Be provocable and contrite
    \item \st{Do not be too clever} Be clever
    \item Adapt to the environment; Adjust to the mean population cooperation
\end{itemize}

The different tournament types as well as the data collection, made
possible due to an open source libary called Axelrod-Python (APL),
are covered in Section~\ref{section:data_collection}. The data set generated
for this work has been made publicly available~\cite{data} and can be used
for further analysis and insights.
Section~\ref{section:top_performances}, focuses on the best performing
strategies for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance, and finally the results are summarised in
Section~\ref{section:conclusion}. This manuscripts uses several parameters,
introduced in the following sections. The full set of
parameters and their definitions are given in Appendix~\ref{app:parameters}.

>>>>

\begin{abstract}
Researchers have explored the performance of Iterated Prisoner's Dilemma strategies
for decades: from the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated learning
structures such as neural networks, many new strategies have been introduced and tested
in a variety of tournaments and population dynamics. Typical results in the literature,
however, rely on performance against a small number of somewhat arbitrarily selected
strategies in a very small number of tournaments, casting doubt on the generalizability
of conclusions. We analyze a large collection of \numberofstrategies typically known
strategies in \numberofalltournaments tournaments, present the top performing strategies across multiple
tournament types, and distill their salient features. The results show that there is not yet a single
strategy that performs well in diverse Iterated Prisoner's Dilemma scenarios.
Nevertheless there are several properties that heavily influence the best performing
strategies, refining the properties described by R. Axelrod in light of
recent and more diverse opponent populations. These are: be nice, be provocable and contrite,
be a little envious, be clever, and adapt to the environment, which includes the parameters
of the tournament (e.g. noise) and the population of opponents. More precisely,
we find that strategies perform best when their probability of cooperation
matches the total tournament population's aggregate cooperation probabilities,
or a proportion thereof in the case of noisy and probabilistically ending tournaments,
and that the manner in which a strategy achieves the ideal cooperation rate is crucial.
\end{abstract}

@@

The Iterated Prisoner's Dilemma (IPD) is a repeated two player game that models
behavioural interactions, specifically interactions where
self-interest clashes with collective interest. At each turn of the game both
players, simultaneously and independently, decide between cooperation (C) and
defection (D), given memory of their prior interactions. The payoffs for each
player, at each turn, is influenced by their own choice and the choice of the
other player. The payoffs of the game are defined by:

@@

where typically \(T > R > P > S\) and \(2R > T + S\). The most common values used in
the literature~\cite{Axelrod1981} are $R=3, P=1, T=5, S=0$. These values are also
used in this work.

Conceptualising strategies and understanding the best way of playing the game
has been of interest to the scientific community since the formulation of the
game in 1950~\cite{Flood1958}. Following the computer tournaments of Axelrod in the
1980's~\cite{Axelrod1980a, Axelrod1980b}, a strategy's performance in a round
robin computer tournament became a common evaluation technique for newly designed
strategies. Many tournaments have followed Axelrod's~\cite{Bendor1991, Stephens2002,
Kendall2007, Stewart2012, Harper2017} and today more hundreds of strategies exist
in the literature.

Axelrod performed two widely-known computer tournaments~\cite{Axelrod1980a,
Axelrod1980b} in the 1980's. The
winner of both tournaments was the simple strategy Tit For Tat (TFT) which cooperated
on the first turn and then simply copied the previous action of its opponent,
retailiating against defections with a defection, and forgiving a defection if followed
by a cooperation. Axelrod concluded that the strategy's robustness was due to four
properties, which he adapted into four suggestions on doing well in an IPD:

\begin{itemize}
    \item Do not be envious by striving for a payoff larger than the opponent's payoff
    \item Be "nice"; Do not be the first to defect
    \item Reciprocate both cooperation and defection; Be provocable to retaliation and forgiveness
    \item Do not be too clever by scheming to exploit the opponent
\end{itemize}

@@

Other works have built upon the limitations of the TFT, and others have shown
that suggestions made by Axelrod did not necessarily lift to alternative environmental rules.
In~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984} it was shown
that TFT suffered in environments with noise. This was
mainly due to the strategy's lack of generosity and contrition. Since TFT immediately
punishes a defection, in a noisy environment it can get stuck in a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced, including
Nice and Forgiving~\cite{Bendor1991}, Generous Tit For Tat~\cite{Nowak1992},
and Pavlov (aka Win Stay Lose Shift) ~\cite{Nowak1993}, as well as later
variants such as OmegaTFT \cite{kendall2007iterated}. Introduction of new strategies
is often accompanied by a claim that the new strategy is the best known despite
only being tested against a small number of opponents, or specific classes of
opponents not necessarily representative of all possible or all published strategies.
The lack of testing against formally defined strategies and tournament winners
is understandable given the effort required to implement
\footnote{Implementing prior strategies faithfully is often extremely difficult or impossible
due to insufficient descriptions and lack of published implementations or code.}
the hundreds of published IPD strategies, yet calls into question any claims of superiority
or robustness of newly introduced
strategies.

A set of envious IPD strategies were introduced called zero-determinant strategies (ZDs)
in~\cite{Press2012}. By
forcing a linear relationship between stationary payoffs ZDs can ensure that they will
never receive less than their memory-one opponents. While ZD strategies
were introduced with a small tournament in which some were reportedly
successful \cite{Stewart2012}, this result has not generally held in future
work. In~\cite{Harper2017} a
series of clever strategies were introduced, and a tournament containing over
200 strategies featured no ZD strategies ranked in top
spots. Instead, the top ranked strategies were a set of ``clever''
(in the sense of Axelrod's characteristics) trained
strategies based on lookup tables~\cite{Axelrod1987}, hidden Markov
models~\cite{Harper2017}, and finite state automata~\cite{Miller1996}.
Similarly, in \cite{mathieu2017}, a set of evolutionarily-trained strategies,
and a pre-selected set of known strategies, outperformed a collection of 12 ZD strategies.

Though only select pieces of work have been discussed, there is a broad collection
of strategies in the literature, and new strategies and competitions are being
published frequently~\cite{Glynatsi2019}. The question, however, still remains
the same: what is the best way to play the game?

Compared to other works, where typically a few selected or introduced strategies
are evaluated on a small number of tournaments and/or small number of opponents,
this manuscript evaluates the performance of \numberofstrategies
strategies in \numberofalltournaments tournaments. Furthermore a large portion
of the strategies used in this manuscript are drawn from the known and named strategies
in IPD literature, including many previous tournament winners,
in contrast to other work that may have randomly generated many essentially arbitrary
strategies (typically restrained to a class such as memory-one strategies,
or those of a certain structural form such as finite state machines or deterministic
memory two strategies). Additionally, our tournaments come in a
number of variations including standard tournaments emulating Axelrod's original tournaments,
tournaments with noise, probabilistic match length, and both noise and probabilistic match length.
This diversity of strategies and tournament types yields new insights and tests
earlier claims in alternative settings against known powerful strategies.

The later part of the paper evaluates
the impact of features on the performance of the strategies using modern
machine learning techniques. These features include measures regarding a
strategy's behaviour and measures regarding the tournaments. The outcomes of our
work reinforce the discussion started by Axelrod, and it concludes that the properties
of a successful strategy in the IPD are:

\begin{itemize}
    \item \st{Do not be envious} Be a little bit envious
    \item Be "nice"; Do not be the first to defect
    \item Reciprocate both cooperation and defection; Be provocable and forgiving
    \item \st{Do not be too clever} It's ok to be clever
    \item Adapt to the environment; Adjust to the mean population cooperation
\end{itemize}

The different tournament types as well as the data collection, made
possible due to an open source library called Axelrod-Python (APL),
are covered in Section~\ref{section:data_collection}. The data set generated
for this work has been made publicly available~\cite{data} and can be used
for further analysis and insights.
Section~\ref{section:top_performances}, focuses on the best performing
strategies for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance, and finally the results are summarised in
Section~\ref{section:conclusion}. This manuscripts uses several parameters,
introduced in the following sections. The full set of
parameters and their definitions are given in Appendix~\ref{app:parameters}.
